---
title: "Effects of resistance training volume on lean body mass and maximal strength"
editor_options: 
  chunk_output_type: console
---

## Introduction

[@krieger2009]

[@rhea2003]

[@Ostrowski1997-at]

[@Cannon2010-un]

[@Berger1962-lc]

[@humburg2007]

## Methods

### Participants

The study included 34 participants between the ages of 18 and 40. To be eligible for the study one had to meat specific health criteria and be a non-smoker. People were excluded if they had any sensitivity to local anesthetics, had participated in resistance training more than once a week in the previous year, had limitations in muscle strength due to injury, or were taking medications that could interfere with training outcomes[@hammarström2020]. Participants are further described in @tbl-participants. Overall, the male participants are older, taller and heavier than the female participants.

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-participants 
#| tbl-cap: "Participant characteristics"



library(tidyverse); library(exscidata);
library(gt)

exscidata::dxadata %>%
  filter(time == "pre",
         include == "incl") %>%
  select(participant, sex, age, height, weight) %>%
  pivot_longer(names_to = "variable",
               values_to = "value",
               cols = age:weight) %>%
  summarise(.by = c(sex, variable),
            m = mean(value),
            s = sd(value),
            n = n()) %>%
  
  mutate(sex = if_else(sex == "male",
                       "Male", "Female"),
         sex = paste0(sex, " n = ", n),
         stat = paste0(round(m, 1),
                       " (",
                       round(s, 1),
                       ")")) %>%
  dplyr::select(sex, variable, stat) %>%
  pivot_wider(names_from = variable,
              values_from = stat) %>%
  gt() %>%
  cols_label(sex = "",
             age = "Age (years)",
             weight = "Body mass (kg)",
             height = "Stature (cm)") %>%
  tab_source_note(
    source_note = "Number of participants are indicated by n. Data are means and standard deviations (SD)."
  )

```

### Study overview

#### Training protocol

#### Assessment of body composition

#### Muscle strength and hypertrophy assesments

### Data analysis

paired t.test

## Results

### Higher training volume results in greater regional hypertrophy

```{r}
#| label: "Load packages and data"
#| warning: false
#| message: false
#| echo: false
#| output: false




library(tidyverse); library(exscidata)
library(lme4)



leanmass <- dxadata %>%
  select(participant:include, lean.left_leg, lean.right_leg) %>%
  pivot_longer(names_to = "leg", 
               values_to = "lean.mass", 
               cols = lean.left_leg:lean.right_leg) %>%
  mutate(leg = if_else(leg == "lean.left_leg", "L", "R"), 
         sets = if_else(multiple == leg, "multiple", "single")) %>%
  select(participant, time, sex, include, sets, leg, lean.mass) %>%
  filter(include == "incl") %>%
 
  mutate(time = factor(time, levels = c("pre", "post")), 
         sets = factor(sets, levels = c("single", "multiple")), 
         leg = paste0(participant, leg)) %>%
  
  print()


leanmass_change_wide <- leanmass %>%
  pivot_wider(names_from = time, values_from = lean.mass) %>%
  mutate(change = post - pre) %>%
  
  select(participant, sex, sets, change) %>%
  pivot_wider(names_from = sets, 
              values_from = change) %>%
  
  mutate(diff = multiple - single) %>%
  
  print()


leanmass_change_wide_time <- leanmass %>%
  pivot_wider(names_from = time, values_from = lean.mass) %>%

  select(participant, sex, sets, pre, post) %>%
  
  print()


tt1 <- with(leanmass_change_wide, t.test(multiple, single, paired = TRUE)) %>%
  print()

str(tt1)

pval <- round(tt1$p.value, 3)

tval <- round(tt1$statistic, 2)

df <- tt1$parameter

cil <- round(tt1$conf.int[1], 1)
ciu <- sprintf("%.1f", round(tt1$conf.int[2], 1))

estimate <- round(tt1$estimate, 1) 

lean.mass.results <- paste0(estimate,
                             " (95% CI: [",
                            cil,
                            ", ",
                            ciu,
                            "], *p*-value = ",
                            pval,
                            ", t<sub>",
                            df, "</sub> = ",
                            tval,")")

```

The mean difference in regional lean body mass change between sets conditions was `r lean.mass.results`.

```{r}
#| echo: false 
#| message: false
#| warning: false
#| label: tbl-change-leanmass 
#| tbl-cap: "Mean lean mass change by sex"
    
library(dplyr)
library(gt)

leanmass_mean_by_sex <- leanmass_change_wide %>%
  group_by(sex) %>%
  summarise(
    mean_single = mean(single, na.rm = TRUE),
    mean_multiple = mean(multiple, na.rm = TRUE),
    mean_diff = mean(diff, na.rm = TRUE)) %>%
  mutate(sex = if_else(sex == "female", "Female", "Male"))

leanmass_mean_by_sex %>%
  gt() %>%
  cols_label(
    sex = "Sex",
    mean_single = "Mean single set change",
    mean_multiple = "Mean multiple set change",
    mean_diff = "Mean difference (multiple - single)") %>%

  fmt_number(
    columns = c(mean_single, mean_multiple, mean_diff),
    decimals = 2) %>%
  cols_align(
    align = "center",
    columns = everything()) %>%
  tab_source_note(
    source_note = "Data is in grams (g)") 
```

@tbl-change-leanmass shows the average lean body mass changes from pre- to post-intervention in male and females. The average change for the single set condition were lower compared to the multiple set condition for both sexes, indicating that multiple sets results in a greater lean mass gain. The last column (Mean difference) indicates that the males had a higher increase of lean mass compared to the female participants.

```{r}
#| echo: false 
#| message: false
#| warning: false
#| label: fig-lean-body-mass 
#| fig-cap: "Mean lean body mass changes from pre- to post-intervention in male and female participants"

leanmass_change_wide1 <- leanmass %>%
  pivot_wider(names_from = time, values_from = lean.mass) %>%
  mutate(change = post - pre) %>%
  select(participant, sex, sets, change) %>%
  pivot_wider(names_from = sets, values_from = change) %>%
  mutate(diff = multiple - single) %>%
  pivot_longer(cols = single:multiple, names_to = "sets", values_to = "change") 


leanmass_summary <- leanmass_change_wide1 %>%
  group_by(sex, sets) %>%
  summarise(
    mean_change = mean(change, na.rm = TRUE),
    se_change = sd(change, na.rm = TRUE) / sqrt(n()),
    .groups = 'drop') %>%
  mutate(sex = if_else(sex == "female", "Female", "Male"),
         sets = if_else(sets == "multiple", "Multiple", "Single"))


ggplot(leanmass_summary, aes(x = sets, y = mean_change, fill = sex)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7), width = 0.6) +
  geom_errorbar(aes(ymin = mean_change - se_change, ymax = mean_change + se_change),
                width = 0.2, position = position_dodge(width = 0.7)) +
  labs(
    x = "Sets",
    y = "Mean Change in Lean Body Mass (g)",
    fill = "Sex"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "top"
  )

```

Comment on @fig-lean-body-mass.

### Higher training volume results in greater muscular strength

```{r}
#| echo: false 
#| message: false
#| warning: false
#| output: false

strengthvolume %>%
  distinct(exercise)
 
 
 

 
str <- strengthvolume %>%
  filter(include == "incl") %>%
  mutate(time = factor(time, levels = c("pre", "session1", 
                                        "week2", "week5", 
                                        "week9", "post"))) 

 

  
 
  str_pre_post <- str %>%
  mutate(time = if_else(time %in% c("pre", "session1"), "pre", time)) %>%
  filter(time %in% c("pre", "post")) %>%   
  group_by(participant, sex, time, sets, exercise, leg) %>%
  summarise(load = max(load, na.rm = TRUE), .groups = "drop") %>%
  filter_all(all_vars(!is.infinite(.)))


  
str_change <- str_pre_post %>% 
pivot_wider(names_from = time, values_from = load) %>%
  mutate(change = post - pre) %>%
  
  select(participant, sex, sets,exercise, change) %>%
  pivot_wider(names_from = sets, 
              values_from = change) %>%
  
  mutate(diff = multiple - single) %>%
  
print()
 
str_tst <-str_change %>% 
  filter(exercise == "legpress") %>% 
  select(participant, sex, single, multiple, diff) %>% 
  print()

tt2 <- t.test(str_tst$multiple, str_tst$single, paired = TRUE)


str(tt2)

pval2 <- round(tt2$p.value, 3)

tval2 <- round(tt2$statistic, 2)

df2 <- tt2$parameter

cil2 <- round(tt2$conf.int[1], 1)
ciu2 <- sprintf("%.1f", round(tt2$conf.int[2], 1))

estimate2 <- round(tt2$estimate, 1) 

strength.results <- paste0(estimate2,
                             " (95% CI: [",
                            cil2,
                            ", ",
                            ciu2,
                            "], *p*-value = ",
                            pval2,
                            ", t<sub>",
                            df2, "</sub> = ",
                            tval2,")")
 
```

The mean difference in maximal strength change between sets conditions was `r strength.results`.

```{r}
#| echo: false 
#| message: false
#| warning: false
#| label: tbl-change-strength 
#| tbl-cap: "Mean muscle strength change by sex"
    
library(dplyr)
library(gt)


strength_mean_by_sex <- str_tst %>%
  group_by(sex) %>%
  summarise(
    mean_single = mean(single, na.rm = TRUE),
    mean_multiple = mean(multiple, na.rm = TRUE),
    mean_diff = mean(diff, na.rm = TRUE)) %>%
  mutate(sex = if_else(sex == "female", "Female", "Male"))


strength_mean_by_sex %>%
  gt() %>%
  cols_label(
    sex = "Sex",
    mean_single = "Mean single set change",
    mean_multiple = "Mean multiple set shange",
    mean_diff = "Mean difference (multiple - single)") %>%
 
  fmt_number(
    columns = c(mean_single, mean_multiple, mean_diff),
    decimals = 2) %>%
  
  cols_align(
    align = "center",
    columns = everything()) %>%
  tab_source_note(
    source_note = "Data is in kilograms (kg)")
 

```

Comment on @tbl-change-strength.

```{r}
#| echo: false 
#| message: false
#| warning: false
#| label: fig-max-strength 
#| fig-cap: "Muscle strength changes from pre- to post-intervention in male and female participants"

str_tst %>%
  pivot_longer(names_to = "sets",
               values_to = "load",
               cols = multiple:single) %>%
  mutate(sets = factor(sets, levels = 
                         c("single", "multiple"))) %>%
  mutate(sex = if_else(sex == "female", "Female", "Male"),
         sets = if_else(sets == "multiple", "Multiple", "Single")) %>%
  ggplot(aes(sets, load, group = participant, 
             color = sex)) +
  geom_line() +
  geom_point() +
  labs(x = "Sets",
       color = "Sex",
       y = "Load (kg)")
```

Comment on @fig-max-strength.
