---
title: "Effects of resistance training volume on lean body mass and maximal strength"
editor_options: 
  chunk_output_type: console
---

## Introduction

Resistance training is a general term referring to exercise requiring one to exert force against a resistance [@kraemer2017]. By using resistance, whether from weights or body weight, resistance training helps improve muscle strength, endurance, and power. This way of training have become a fundamental element in fitness and rehabilitation due to its variety of health benefits, as it enhances physical performance and improves overall quality of life. Such benefits are valuable across many fields, from sports to general health and longevity. Investigating the effects of resistance training is therefore important for understanding how to best utilize its health benefits. Different variables in resistance training, such as volume, may affect the outcome, making it essential to examine how specific training protocols contribute to muscular strength and body composition.

Resistance training volume refers to the total amount of work performed in a single resistance training session or across multiple sessions [@Ostrowski1997-at]. Volume is a crucial variable in resistance training as it influences the training adaptions, such as muscle growth, lean body mass and muscle strength. Muscle growth, or hypertrophy, involves an increase in the number of muscle fibers and thus increased muscular size while lean body mass refers to the total body mass without the weight of body fat [@kraemer2017]. Muscle hypertrophy is one component of lean mass gain as gains in lean mass are often attributed to hypertrophy. Muscle strength refers to the maximum force or tension that a muscle or group of muscles can generate during a single maximal effort and is influenced by factors such as hypertrophy [@kraemer2004]. Finding the right volume balance is essential as different volumes may lead to different results. Some authors have claimed that multiple sets are necessary to optimize strength gains, while others have argued that a single set per exercise is all that is necessary, and further gains are not achieved by successive sets [@krieger2009]. Ostrowski et al. investigated the impact of different volumes of resistance training on muscle size and function over a 10-week period and found that regardless of the volume of training, all groups experienced increases in muscle size and strength [@Ostrowski1997-at]. These findings are supported by Cannon & Marino as their findings demonstrate that additional neuromuscular adaption during early-phase moderate-intensity resistance training may not be elicited through higher-volume training when training loads are matched provided that a minimal volume threshold is attained [@Cannon2010-un]. However, evidence from several other studies supports greater strength gains with multiple sets. One of the most notable scientific studies was performed in the early 1960s and found that different training volumes and intensities elicit different magnitudes of strength gains as three sets of six repetitions resulted in the greatest strength increases [@rhea2003] & [@Berger1962-lc]). A more recent study, done by Humburg et al., support these findings as they argue that improvements for 1RM is significantly higher during a 3-set program compared to a 1-set program [@humburg2007].

As a result, there has been considerable debate over the optimal number of sets per exercise to improve musculoskeletal strength during a resistance exercise program [@krieger2009]. Additionally, much research has been performed examining strength increases with training, but they provide little insight into the magnitude of strength gains along the continuum of training volume [@rhea2003]. Hence, the purpose of this study is to identify the effect of single- and multiple-set programs on lean body mass and maximal strength by using a within-participant study design.

## Methods

### Participants

The study included 34 participants between the ages of 18 and 40. To be eligible for the study one had to meat specific health criteria and be a non-smoker. People were excluded if they had any sensitivity to local anesthetics, had participated in resistance training more than once a week in the previous year, had limitations in muscle strength due to injury, or were taking medications that could interfere with training outcomes [@hammarström2020]. Participants are further described in @tbl-participants. Overall, the male participants are older, taller and heavier than the female participants.

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-participants 
#| tbl-cap: "Participant characteristics"



library(tidyverse); library(exscidata);
library(gt)

exscidata::dxadata %>%
  filter(time == "pre",
         include == "incl") %>%
  select(participant, sex, age, height, weight) %>%
  pivot_longer(names_to = "variable",
               values_to = "value",
               cols = age:weight) %>%
  summarise(.by = c(sex, variable),
            m = mean(value),
            s = sd(value),
            n = n()) %>%
  
  mutate(sex = if_else(sex == "male",
                       "Male", "Female"),
         sex = paste0(sex, " n = ", n),
         stat = paste0(round(m, 1),
                       " (",
                       round(s, 1),
                       ")")) %>%
  dplyr::select(sex, variable, stat) %>%
  pivot_wider(names_from = variable,
              values_from = stat) %>%
  gt() %>%
  cols_label(sex = "",
             age = "Age (years)",
             weight = "Body mass (kg)",
             height = "Stature (cm)") %>%
  tab_source_note(
    source_note = "Number of participants are indicated by n. Data are means and standard deviations (SD).") %>%
  tab_options(
    table.font.size = px(9), 
    table.width = px(500)    
  )

```

### Study overview

The 12-week training intervention consisted of full-body resistance exercises. Leg exercises were done one leg at a time, allowing for the comparison of different training volumes and its effect on each participant. One leg was randomly assigned to perform a single set of each exercise, while the other leg performed three sets. This allowed each participant to engage in both workout protocols. Muscle strength was evaluated at the beginning of the intervention (pre), and at the end of the 12 weeks (post). Body composition, including measurements of hypertrophy and lean mass gain, were taken at the start and conclusion of the study.

#### Training protocol

Participants completed a standardized warm-up before each workout. The warm-up included cycling on an ergometer for 5 minutes with a rate of perceived exertion (RPE) score of 12-14 as the intensity should be moderate. Secondly, they did bodyweight exercises such as back extensions, push-ups and squats for 10 repetitions. Lastly, participants performed one set of 10 reps at 50% of their one-repetition maximum (1RM) for each exercise. Leg resistance exercises were performed as either one set (single) or three sets (multiple), with the single set completed between the second and third set in the multiple-set routine. Exercises were performed in order with unilateral leg press first, then leg curl and lastly knee extension. Rest intervals between the sets ranged from 90 to 180 seconds. The training intensity gradually increased over the course of the intervention. The participants started with 10RM for the first two weeks, then 8RM for the following three weeks, and lastly 7RM for the remaining seven weeks. One session a week with reduced loads (90%) were introduced from the ninth session. High-intensity training sessions were spaced at least 48 hours apart, while sessions with reduced loads were scheduled at least 24 hours apart from other sessions.

#### Maximal strength and lean mass assessments

Maximal strength for unilateral leg press was evaluated using the one-repetition maximum (1RM) method. 1RM was found by progressively increasing the weight until the participant could no longer complete a full range of motion. The successful lift with the highest load was recorded as the 1RM. Lean mass (and hypertrophy) was measured regionally and determined before and after the training intervention using a dual-energy X-ray absorptiometry (DXA) scanner and magnetic resonance imaging (MRI). MRI specifically determined the knee-extensor muscle cross-sectional area (CSA; vastus lateralis, medialis, intermedius and rectus femoris).

### Data analysis

In this study, we used a paired t-test to compare the pre- and post-intervention measurements between single and multiple sets for each participant. The data was analyzed in RStudio.

## Results

### Higher training volume results in greater regional hypertrophy

```{r}
#| label: "Load packages and data"
#| warning: false
#| message: false
#| echo: false
#| output: false




library(tidyverse); library(exscidata)
library(lme4)



leanmass <- dxadata %>%
  select(participant:include, lean.left_leg, lean.right_leg) %>%
  pivot_longer(names_to = "leg", 
               values_to = "lean.mass", 
               cols = lean.left_leg:lean.right_leg) %>%
  mutate(leg = if_else(leg == "lean.left_leg", "L", "R"), 
         sets = if_else(multiple == leg, "multiple", "single")) %>%
  select(participant, time, sex, include, sets, leg, lean.mass) %>%
  filter(include == "incl") %>%
 
  mutate(time = factor(time, levels = c("pre", "post")), 
         sets = factor(sets, levels = c("single", "multiple")), 
         leg = paste0(participant, leg)) %>%
  
  print()


leanmass_change_wide <- leanmass %>%
  pivot_wider(names_from = time, values_from = lean.mass) %>%
  mutate(change = post - pre) %>%
  
  select(participant, sex, sets, change) %>%
  pivot_wider(names_from = sets, 
              values_from = change) %>%
  
  mutate(diff = multiple - single) %>%
  
  print()


leanmass_change_wide_time <- leanmass %>%
  pivot_wider(names_from = time, values_from = lean.mass) %>%

  select(participant, sex, sets, pre, post) %>%
  
  print()


tt1 <- with(leanmass_change_wide, t.test(multiple, single, paired = TRUE)) %>%
  print()

str(tt1)

pval <- round(tt1$p.value, 3)

tval <- round(tt1$statistic, 2)

df <- tt1$parameter

cil <- round(tt1$conf.int[1], 1)
ciu <- sprintf("%.1f", round(tt1$conf.int[2], 1))

estimate <- round(tt1$estimate, 1) 

lean.mass.results <- paste0(estimate,
                             " (95% CI: [",
                            cil,
                            ", ",
                            ciu,
                            "], *p*-value = ",
                            pval,
                            ", t<sub>",
                            df, "</sub> = ",
                            tval,")")

```

The mean difference in grams (g) for regional lean body mass change between single and multiple sets was `r lean.mass.results`.

```{r}
#| echo: false 
#| message: false
#| warning: false
#| label: tbl-change-leanmass 
#| tbl-cap: "Mean lean mass change by sex"
    
library(dplyr)
library(gt)

leanmass_mean_by_sex <- leanmass_change_wide %>%
  group_by(sex) %>%
  summarise(
    mean_single = mean(single, na.rm = TRUE),
    mean_multiple = mean(multiple, na.rm = TRUE),
    mean_diff = mean(diff, na.rm = TRUE)) %>%
  mutate(sex = if_else(sex == "female", "Female", "Male"))

leanmass_mean_by_sex %>%
  gt() %>%
  cols_label(
    sex = "Sex",
    mean_single = "Mean single set change",
    mean_multiple = "Mean multiple set change",
    mean_diff = "Mean difference (multiple - single)") %>%

  fmt_number(
    columns = c(mean_single, mean_multiple, mean_diff),
    decimals = 2) %>%
  cols_align(
    align = "center",
    columns = everything()) %>%
  tab_source_note(
    source_note = "Data is in grams (g) and shows the average lean body mass changes from pre- to post-intervention in male and females. The average change for the single set condition where lower compared to the multiple set condition for both sexes, indicating that multiple sets results in a greater lean mass gain. The last column (Mean difference) indicates that the males had a higher increase of lean mass between the different sets compared to the female participants") %>%
  tab_options(
    table.font.size = px(9), 
    table.width = px(500)    
  )
```

```{r}
#| echo: false 
#| message: false
#| warning: false
#| label: fig-lean-body-mass 
#| fig-cap: "Mean lean body mass changes from pre- to post-intervention in male and female participants"

leanmass_change_wide1 <- leanmass %>%
  pivot_wider(names_from = time, values_from = lean.mass) %>%
  mutate(change = post - pre) %>%
  select(participant, sex, sets, change) %>%
  pivot_wider(names_from = sets, values_from = change) %>%
  mutate(diff = multiple - single) %>%
  pivot_longer(cols = single:multiple, names_to = "sets", values_to = "change") 


leanmass_summary <- leanmass_change_wide1 %>%
  group_by(sex, sets) %>%
  summarise(
    mean_change = mean(change, na.rm = TRUE),
    se_change = sd(change, na.rm = TRUE) / sqrt(n()),
    .groups = 'drop') %>%
  mutate(sex = if_else(sex == "female", "Female", "Male"),
         sets = if_else(sets == "multiple", "Multiple", "Single"))


ggplot(leanmass_summary, aes(x = sets, y = mean_change, fill = sex)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7), width = 0.6) +
  geom_errorbar(aes(ymin = mean_change - se_change, ymax = mean_change + se_change),
                width = 0.2, position = position_dodge(width = 0.7)) +
  labs(
    x = "Sets",
    y = "Mean Change in Lean Body Mass (g)",
    fill = "Sex",
    caption = "Multiple sets give a higher average increase in lean mass gains compared to a single set in both female and male \nparticipants. The black lines represent the standard error of the mean (SE) for each condition (single and \nmultiple sets) and for each sex (male and female), indicating the variability of the mean change values. "
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "top",
    plot.caption = element_text(hjust = 0, size = 7, face = "italic")
  )
 

```

### Higher training volume results in greater muscular strength

```{r}
#| echo: false 
#| message: false
#| warning: false
#| output: false

strengthvolume %>%
  distinct(exercise)
 
 
 

 
str <- strengthvolume %>%
  filter(include == "incl") %>%
  mutate(time = factor(time, levels = c("pre", "session1", 
                                        "week2", "week5", 
                                        "week9", "post"))) 

 

  
 
  str_pre_post <- str %>%
  mutate(time = if_else(time %in% c("pre", "session1"), "pre", time)) %>%
  filter(time %in% c("pre", "post")) %>%   
  group_by(participant, sex, time, sets, exercise, leg) %>%
  summarise(load = max(load, na.rm = TRUE), .groups = "drop") %>%
  filter_all(all_vars(!is.infinite(.)))


  
str_change <- str_pre_post %>% 
pivot_wider(names_from = time, values_from = load) %>%
  mutate(change = post - pre) %>%
  
  select(participant, sex, sets,exercise, change) %>%
  pivot_wider(names_from = sets, 
              values_from = change) %>%
  
  mutate(diff = multiple - single) %>%
  
print()
 
str_tst <-str_change %>% 
  filter(exercise == "legpress") %>% 
  select(participant, sex, single, multiple, diff) %>% 
  print()

tt2 <- t.test(str_tst$multiple, str_tst$single, paired = TRUE)


str(tt2)

pval2 <- round(tt2$p.value, 3)

tval2 <- round(tt2$statistic, 2)

df2 <- tt2$parameter

cil2 <- round(tt2$conf.int[1], 1)
ciu2 <- sprintf("%.1f", round(tt2$conf.int[2], 1))

estimate2 <- round(tt2$estimate, 1) 

strength.results <- paste0(estimate2,
                             " (95% CI: [",
                            cil2,
                            ", ",
                            ciu2,
                            "], *p*-value = ",
                            pval2,
                            ", t<sub>",
                            df2, "</sub> = ",
                            tval2,")")
 
```

The mean difference in kilograms (kg) for maximal strength change between single and multiple sets was `r strength.results`.

```{r}
#| echo: false 
#| message: false
#| warning: false
#| label: tbl-change-strength 
#| tbl-cap: "Mean muscle strength change by sex"

    
library(dplyr)
library(gt)


strength_mean_by_sex <- str_tst %>%
  group_by(sex) %>%
  summarise(
    mean_single = mean(single, na.rm = TRUE),
    mean_multiple = mean(multiple, na.rm = TRUE),
    mean_diff = mean(diff, na.rm = TRUE)) %>%
  mutate(sex = if_else(sex == "female", "Female", "Male"))


strength_mean_by_sex %>%
  gt() %>%
  cols_label(
    sex = "Sex",
    mean_single = "Mean single set change",
    mean_multiple = "Mean multiple set shange",
    mean_diff = "Mean difference (multiple - single)") %>%
 
  fmt_number(
    columns = c(mean_single, mean_multiple, mean_diff),
    decimals = 2) %>%
  
  cols_align(
    align = "center",
    columns = everything()) %>%
  tab_source_note(
    source_note = "Data is in kilograms (kg) and shows the average muscle strength changes from pre- to post-intervention in male and females. The average change for the single set condition where lower compared to the multiple set condition for both sexes, indicating that multiple sets results in a greater strength gain. The last column (Mean difference) indicates that the females had a greater difference in gain of muscle strength between the sets compared to the male participants.") %>%
  tab_options(
    table.font.size = px(9), 
    table.width = px(500)    
  )
 

```

```{r}
#| echo: false 
#| message: false
#| warning: false
#| label: fig-max-strength 
#| fig-cap: "Muscle strength changes from pre- to post-intervention in male and female participants"

strength_summary <- str_change %>%
  filter(exercise == "legpress") %>%
  pivot_longer(cols = single:multiple, names_to = "sets", values_to = "change") %>%
  group_by(sex, sets) %>%
  summarise(
    mean_change = mean(change, na.rm = TRUE),
    se_change = sd(change, na.rm = TRUE) / sqrt(n()),  
    .groups = 'drop')

ggplot(strength_summary, aes(x = sets, y = mean_change, fill = sex)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7), width = 0.6) +
  geom_errorbar(aes(ymin = mean_change - se_change, ymax = mean_change + se_change),
                width = 0.2, position = position_dodge(width = 0.7)) +
  labs(
    x = "Sets",
    y = "Mean Change in Strength (kg)",
    fill = "Sex",
    caption = "Multiple sets give a higher average increase in maximal strength compared to a single set in both female and male \nparticipants. The black lines represent the standard error of the mean (SE) for each condition (single and \nmultiple sets) and for each sex (male and female), indicating the variability of the mean change values."
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "top",
    plot.caption = element_text(hjust = 0, size = 7, face = "italic"))

```

## Discussion

The paired t-test results for the lean mass and maximal strength data indicate that the difference in change between the single-set and multiple-set conditions was statistically significant, with a mean difference of `r lean.mass.results` and `r strength.results` respectively. The p-value of `r pval` and `r pval2` suggests that this result is unlikely to be due to random chance, as it falls below the significance threshold of 0.05. Furthermore, the 95% confidence interval (CI) demonstrates that the true mean difference in regional lean body mass gain lies between `r cil` grams and `r ciu` grams, while the true mean difference in strength lies between `r cil2` kilograms and `r ciu2` kilograms. These findings suggest that the observed effect is consistent and meaningful, but also highlighting some variability across participants.

Overall, the results indicate that both training volumes were effective in improving muscle function, with increased gains in lean body mass and maximal strength. Improvements from pre- to post-intervention where on average higher for the leg performing multiple sets of the exercise compared to the leg who did a single set. Both male and female participants demonstrated similar patterns of improvement, however, males experienced greater gains in lean mass with multiple sets (see @tbl-change-leanmass), while females showed more pronounced increases in maximal strength (see @tbl-change-strength). All in all, these findings suggest that a moderate-volume training regimen is more effective than a low-volume regimen in enhancing lean mass and maximal strength in both men and women.

Our findings align with a growing body of research that indicates higher training volumes generally lead to more significant improvements in both lean mass and muscle strength. Rhea et al. [@rhea2003], similarly reported that multiple sets resulted in superior strength gains compared to single sets. Additionally, research by Berger and Humburg et al. suggests that training volume plays a significant role in optimizing strength outcomes, particularly with multi-set protocols [@Berger1962-lc];[@humburg2007]. On the other hand, it is worth nothing that regardless of training volume, we saw that on average both sexes exhibited increases in muscle size and strength. These findings align with Ostrowski’s notion that resistance training is beneficial across various volumes [@Ostrowski1997-at]. However, our results challenge his conclusion that there are no significant differences between different training volumes, suggesting that a higher volume of training may play a more significant role in optimizing muscle growth and strength development. Moreover, we found that the magnitude of the improvement in lean mass was more pronounced in male participants, whereas females showed more significant strength gains. These findings may align with those of Cannon and Marino who suggested that training volume and intensity have different impacts depending on the individual’s training status and the specific adaptions targeted [@Cannon2010-un].

## Conclusion

In conclusion, the results demonstrate that performing multiple sets leads to greater improvements in both lean mass gains and muscular strength compared to a single set. These findings are in line with previous findings [@rhea2003], [@Berger1962-lc] and [@humburg2007] while challenging the conclusion of studies who suggest that there are no differences in outcomes between training volumes [@Ostrowski1997-at]. In other words, our study highlights the importance of training volumes in optimizing muscular adaptions as our results indicate that increasing the training volume may be more effective for both muscle size and strength gains. However, further research is needed to explore the underlying mechanisms for the differential effects of resistance training volume.
